---
title: "Conversion from OR to RR or RD"
author: "Ehsan Karim: ehsank.com"
date: "`r format(Sys.time(), '%d %B %Y')`"
always_allow_html: yes
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    toc: yes
    number_sections: true
    toc_depth: 3
    toc_float: 
      collapsed: true
      smooth_scroll: true
    theme: lumen
    highlight: textmate
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
  slidy_presentation:
    toc: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(epiDisplay)
require(sjstats)
```


# Notations

- OR = Odds ratio
- RR = Risk ratio
- RD = Risk difference
- p0 = Background event rate (e.g., for unexposed group)
- p1 = Risk for experiencing event for exposed group
- O0 = Background odds of experiencing events
- O1 = Odds for experiencing event for exposed group

# Functions

## OR to RD

Following ref 1
```{r}
OR2RD <- function(OR,p0){
  O0 = p0/(1-p0)
  RD = O0*(OR-1)/( (1+OR*O0)*(1+O0) )
  return(RD)
}
```

## OR to RR

Following ref 1
```{r}
OR2RR <- function(OR,p0){
  O0 = p0/(1-p0)
  RD = O0*(OR-1)/( (1+OR*O0)*(1+O0) )
  RR = (RD+p0)/p0
  return(RR)
}
```

Following ref 2
```{r}
OR2RRx <- function(OR,p0){
  O0 = p0/(1-p0)
  O1 = OR*O0 
  p1 = O1/(1+O1) 
  RR = p1/p0
  return(RR)
}
```

Following ref 3
```{r}
OR2RRy <- function(OR,p0){
  RR = OR/(1 - p0 + (OR * p0))
  return(RR)
}
```

# Testing the Function

## OR to RD

### Option 1

Using derivation by a student: Liang Xu.

```{r}
OR = seq(0.1,2,0.1)
p0.list = seq(0.1,.9,0.1)
plot(OR,seq(-0.55,0.15,length=20), type = "n", xlab = "OR", ylab = "RD")
for (i in p0.list){
  p0 = i 
  p1 = OR * (p0/(1 - p0))/(1 + OR * (p0/(1 - p0)))
  RD = p1-p0
  lines(OR,RD, col = i*10, lty = i*10)
}
legend("bottomright", legend=p0.list, title="p0",
       col=p0.list*10, lty=p0.list*10, cex=0.8)
```


### Option 2

```{r}
OR = seq(0.1,2,0.1)
p0.list = seq(0.1,.9,0.1)
plot(OR,seq(-0.55,0.15,length=20), type = "n", xlab = "OR", ylab = "RD")
for (i in p0.list){
  RD = OR2RD(OR = OR, p0 = i)
  lines(OR,RD, col = i*10, lty = i*10)
}
legend("bottomright", legend=p0.list, title="p0",
       col=p0.list*10, lty=p0.list*10, cex=0.8)
```


## OR to RR

### Option 1

```{r}
OR = seq(0.1,2,0.1)
p0.list = seq(0.1,.9,0.1)
plot(OR,seq(0,2,length=20), type = "n", xlab = "OR", ylab = "RR")
for (i in p0.list){
  p0 = i
  p1 = OR * (p0/(1 - p0))/(1 + OR * (p0/(1 - p0)))
  RR = p1/p0
  lines(OR,RR, col = i*10, lty = i*10)
}
legend("bottomright", legend=p0.list, title="p0",
       col=p0.list*10, lty=p0.list*10, cex=0.8)
```


### Option 2

```{r}
OR = seq(0.1,2,0.1)
p0.list = seq(0.1,.9,0.1)
plot(OR,seq(0,2,length=20), type = "n", xlab = "OR", ylab = "RR")
for (i in p0.list){
  RR = OR2RR(OR = OR, p0 = i)
  lines(OR,RR, col = i*10, lty = i*10)
}
legend("bottomright", legend=p0.list, title="p0",
       col=p0.list*10, lty=p0.list*10, cex=0.8)
```


### Option 3

```{r}
OR = seq(0.1,2,0.1)
p0.list = seq(0.1,.9,0.1)
plot(OR,seq(0,2,length=20), type = "n", xlab = "OR", ylab = "RR")
for (i in p0.list){
  RR = OR2RRx(OR = OR, p0 = i)
  lines(OR,RR, col = i*10, lty = i*10)
}
legend("bottomright", legend=p0.list, title="p0",
       col=p0.list*10, lty=p0.list*10, cex=0.8)
```

# Working with a data

## Titanic data

### OR

```{r}
url <- "http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/titanic.txt"
titanic <- read.csv(file = url, stringsAsFactors = FALSE)
titanic$age[is.na(titanic$age)] <- median(titanic$age, na.rm = TRUE) 
fit.OR <- glm(survived ~ sex,
                         family = binomial("logit"), data = titanic)
est.OR = exp(coef(fit.OR))
est.OR
require(epiDisplay)
logistic.display(fit.OR)
```

### RD

- The outcome variable is `survived`
- `sex` variable is considered as the exposure variable: `Female` categoty is the reference category, and hence considered here as the background population to calculate $p0$.
- Stating values were identified by grid search (e.g., using `for` loop within plausible values)

```{r}
tablex <- table(titanic$sex,titanic$survived)
tablex
survival.rate = tablex[,"1"]/sum(tablex[,"1"])
survival.rate
p0 <- survival.rate[1]
p0
OR2RD(OR=est.OR[2],p0=p0)
fit.RD <- glm(survived ~ sex, start = c(.1,-.01),
                         family = binomial("identity"), data = titanic)
fit.RD
```

### RR

```{r}
OR2RR(OR=est.OR[2],p0=p0)
OR2RRx(OR=est.OR[2],p0=p0)
OR2RRy(OR=est.OR[2],p0=p0)
require(sjstats)
or_to_rr(or=est.OR, p0=p0)
```

- Stating values were identified by grid search (e.g., using `for` loop within plausible values)

```{r}
fit.RR <- glm(survived ~ sex, start=c(-1,.5),
                         family = binomial("log"), data = titanic)
fit.RR
exp(coef(fit.RR))
```

# References

1. [Popham, F. (2016). Converting between marginal effect measures from binomial models. International journal of epidemiology, 45(2), 590-591.](https://academic.oup.com/ije/article/45/2/590/2572549)
2. [StatsToDo : Odd and Risk Interconversion Explained](https://www.statstodo.com/OddRiskConversion_Exp.php)
3. [Get relative risks estimates from logistic regressions or odds ratio values](https://strengejacke.github.io/sjstats/reference/odds_to_rr.html)